---
title: "PT/FT Group Differences Analyses - Moderation"
author: "Virginia Marchman"
date: "May 6, 2021; updated 6/3/2021, updated 1/13/2022; updated 5/25/2022; updated 8/20/2022"
output:
  html_document: 
   toc: true
   toc_float: true
---

```{r, echo = F}
library(knitr)
opts_chunk$set(echo=TRUE, 
               warning=FALSE, message=FALSE, 
               cache=FALSE)
options(width = 100)
```

# This is the code to generate the analyses in Marchman, V.A., Ashland, M., Loi, E.C., Shannon, K. A., Munevar, M., Fernald, A, & Feldman, H. M. Associations between early efficiency in language processing and learning outcomes in children born full term and preterm: Similarities and differences.

Load libraries and set theme
```{r}
library(tidyverse)
library(effsize)
library(stargazer)
library(psych)
library(powerAnalysis)
library(reghelper)

theme_set(theme_bw())
```


# Data prepping
### Load data
```{r}

PTFT <- read_csv("PTFTdata.csv")

PTFT <- as.data.frame(PTFT)
```



### Check data
```{r}
# Sex - 46 females, 49 males

PTFT %>% 
  group_by(Group) %>% 
  count()

PTFT %>% 
  group_by(Group, Sex) %>% 
  count()
```

### Transform variables (e.g., character to factor)
```{r}
PTFT <- PTFT %>% 
  mutate(SubjectID = factor(SubjectID),
         Sex = factor(Sex), Group = factor(Group))
```

# Demographic variables
### Descriptive statistics - t-tests by Birth Group - Table 1

```{r}

t.test(GADays ~ Group, data = PTFT, var.equal = FALSE)
t.test(BirthWtGm ~ Group, data = PTFT, var.equal = FALSE)
t.test(MomEdu ~ Group, data = PTFT, var.equal = FALSE)
t.test(HI_SES ~ Group, data = PTFT, var.equal = FALSE)
t.test(AgeAtFirstVisit_PDEV18ATL318C ~ Group, data = PTFT, var.equal = FALSE)
t.test(AgeAtFirstVisit_PDEV54CTL354C ~ Group, data = PTFT, var.equal = FALSE)

```


### Percentage of children by birth order
```{r}
PTFT %>% 
  group_by(BirthOrder, Group) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(percentage = n/sum(n))

PTFTbirthgp <- PTFT %>%
  mutate(
    birthgroup = case_when(
          BirthOrder == 1 ~ 1,
          BirthOrder == 2 ~ 2,
          BirthOrder > 2 ~ 3)
    )

PTFTbirthgp %>% 
  group_by(birthgroup, Group) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(percentage = n/sum(n))

```
### Number of trials with known words by birthgroup.
```{r}


PTFT %>% group_by(Group) %>%
  summarise(mean_Known = mean(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T),
            sd_Known = sd(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T),
            min_Known = min(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T),
            max_Known = max(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T))

t.test(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A ~ Group, data = PTFT)

```

### Number of trials for RT and test for differences between birthgroup.
```{r}

PTFT %>% group_by(Group) %>%
  summarise(mean_trials = mean(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            sd_trials = sd(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T),
            min_trials = min(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            max_trials = max(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T))

t.test(DRTAB183001800knownnumPDEV18ATL318C ~ Group, data = PTFT)

```



### Number of trials for Accuracy and test for differences between birthgroups.
```{r}

PTFT %>% group_by(Group) %>%
  summarise(mean_trials = mean(AccAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            sd_trials = sd(AccAB183001800knownnumPDEV18ATL318C, na.rm = T),
            min_trials = min(AccAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            max_trials = max(AccAB183001800knownnumPDEV18ATL318C, na.rm = T))

t.test(AccAB183001800knownnumPDEV18ATL318C ~ Group, data = PTFT)

```



# LWL measures at 18 months - Table 2
```{r}
# Accuracy - t-test against chance overall

t.test(PTFT$AccAB183001800knownPDEV18ATL318C, y = NULL, mu = .50, var.equal = FALSE)

(mean(PTFT$AccAB183001800knownPDEV18ATL318C) - .50) / sd(PTFT$AccAB183001800knownPDEV18ATL318C)

# Accuracy - t-test by group

t.test(AccAB183001800knownPDEV18ATL318C ~ Group, data = PTFT)

# RT - t-test by group

t.test(RT18A18C_NO1RT ~ Group, data = PTFT)

# Correlations between Accuracy and RT

cor.test(PTFT$AccAB183001800knownPDEV18ATL318C, PTFT$RT18A18C_NO1RT)

```
# LWL measures at 18 months - within PT and FT group separately
```{r}

# Accuracy - t-test against chance - PT group only

PTonly <- PTFT %>%
  filter(Group == "PT")

  t.test(PTonly$AccAB183001800knownPDEV18ATL318C, y = NULL, mu = .50, var.equal = FALSE)

# Accuracy - t-test against chance - FT group only

FTonly <- PTFT %>%
  filter(Group == "FT")

  t.test(FTonly$AccAB183001800knownPDEV18ATL318C, y = NULL, mu = .50, var.equal = FALSE)

```

# Child language measures at 4.5 years - Table 2
# Group difference by sex
```{r}

PTFT %>% group_by(Sex) %>%
  summarise(mean_PPVT = mean(PPVT_RecVocab_standard, na.rm = T), 
            sd_PPVT = sd(PPVT_RecVocab_standard, na.rm = T),
            mean_CELF = mean(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard, na.rm = T), 
            se_CELF = sd(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard, na.rm = T),
            mean_IQ = mean(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor, na.rm = T), 
            se_IQ = sd(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor, na.rm = T))


# PPVT
t.test(PPVT_RecVocab_standard ~ Sex, data = PTFT)

# CELF
t.test(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ Sex, data = PTFT)

#IQ

t.test(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ Sex, data = PTFT)

```
#Group differences by Group
```{r}

PTFT %>% group_by(Group) %>%
  summarise(mean_PPVT = mean(PPVT_RecVocab_standard, na.rm = T), 
            sd_PPVT = sd(PPVT_RecVocab_standard, na.rm = T),
            mean_CELF = mean(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard, na.rm = T), 
            se_CELF = sd(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard, na.rm = T),
            mean_IQ = mean(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor, na.rm = T), 
            se_IQ = sd(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor, na.rm = T))

# PPVT
t.test(PPVT_RecVocab_standard ~ Group, data = PTFT)

# CELF
t.test(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ Group, data = PTFT)

#IQ

t.test(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ Group, data = PTFT)

```

# Regression models
### Models 1a to 1e PPVT Receptive Vocabulary - Table 3

```{r}

# covariates SES and Age and Group
m1a <- lm(PPVT_RecVocab_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group, data = PTFT)

# covariates + Group + Accuracy
m1b <- lm(PPVT_RecVocab_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group + AccAB183001800knownPDEV18ATL318C, data = PTFT)

# covariates + Group + Accuracy + Acc x Group
m1c <- lm(PPVT_RecVocab_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group * AccAB183001800knownPDEV18ATL318C, data = PTFT)

# covariates + Group + RT
m1d <- lm(PPVT_RecVocab_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group + RT18A18C_NO1RT, data = PTFT)

# covariates + Group + RT + RT x Group
m1e <- lm(PPVT_RecVocab_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group * RT18A18C_NO1RT, data = PTFT)

# comparisons
anova(m1a, m1b)
anova(m1b, m1c)
anova(m1a, m1d)
anova(m1d, m1e)

stargazer(m1a, m1b, m1c, m1d, m1e, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 3,
         dep.var.labels = c("Receptive Vocabulary"),
         covariate.labels=c("Age at test", "SES",
                            "Group",
                            "Acc",
                            "Acc x Group",
                            "RT",
                            "RT x Group"))

```

### Models 2a to 2e CELF Expressive Language - Table 4
```{r}
# covariates SES, Age, Sex and Group
m2a <- lm(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Sex + Group, data = PTFT)

# covariates + Group + Accuracy
m2b <- lm(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Sex + Group + AccAB183001800knownPDEV18ATL318C, data = PTFT)

# covariates + Group + Accuracy + Acc x Group
m2c <- lm(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Sex + Group * AccAB183001800knownPDEV18ATL318C, data = PTFT)

# covariates + Group + RT
m2d <- lm(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Sex + Group + RT18A18C_NO1RT, data = PTFT)

# covariates + Group + RT + RT x Group
m2e <- lm(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Sex + Group * RT18A18C_NO1RT, data = PTFT)

# comparisons
anova(m2a, m2b)
anova(m2b, m2c)
anova(m2a, m2d)
anova(m2d, m2e)

stargazer(m2a, m2b, m2c, m2d, m2e, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 3,
         dep.var.labels = c("Expressive Language"),
         covariate.labels=c("Age at test", "SES", "Sex",
                            "Group",
                            "Acc",
                            "Acc x Group",
                            "RT",
                            "RT x Group"))

```
### ### Models 3a to 3e - Non Verbal IQ - Table 5
```{r}
# covariates SES and Age and Group

m3a <- lm(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group, data = PTFT)

# covariates + Group + Accuracy
m3b <- lm(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group + AccAB183001800knownPDEV18ATL318C, data = PTFT)

# covariates + Group + Accuracy + Acc x Group
m3c <- lm(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group * AccAB183001800knownPDEV18ATL318C, data = PTFT)

# covariates + Group + RT
m3d <- lm(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group + RT18A18C_NO1RT, data = PTFT)

# covariates + Group + RT + RT x Group
m3e <- lm(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group * RT18A18C_NO1RT, data = PTFT)

# Adding CELF as an additional control
m3f <- lm(IQPDEV54TL354_Leiter_BriefIQ_IQ_compscor ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard + Group * RT18A18C_NO1RT, data = PTFT)

# comparisons
anova(m3a, m3b)
anova(m3b, m3c)
anova(m3a, m3d)
anova(m3d, m3e)
anova(m3d, m3f)


stargazer(m3a, m3b, m3c, m3d, m3e, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 2,
         dep.var.labels = c("Non-verbal IQ"),
         covariate.labels=c("Age at test", "SES",
                            "Group",
                            "Acc",
                            "Acc x Group",
                            "RT",
                            "RT x Group"))


# Getting simple slopes for the interaction between group and RT

simple_slopes(m3e)


```
